### 2022-06-27 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@3f97d39ccff49b8819786ddb750ed9dd1c424697 🚀

\hypertarget{host-richness-does-not-predict-virus-distinctiveness}{%
\subsection{Host richness does not predict virus
distinctiveness}\label{host-richness-does-not-predict-virus-distinctiveness}}

Bats, the second most diverse groups of mammals, are found worldwide;
gradients in bat species richness generally track broader patterns of
mammal diversity, with a striking Neotropical hotspot (especially in the
Amazon basin) and a secondary hotspot centered in the southeast Asian
peninsula. These hotspots of bat diversity are generally presumed to be
hotspots of viral diversity, and therefore areas of concern for human
health.\textsuperscript{2,13} However, the hotspots of bat
betacoronavirus reservoirs show a distinct pattern, with primary
hotspots (both in terms of size and higher values) of host richness
situated primarily South-Eastern Asia, parts of Southern Europe, and to
a lesser extent parts of Africa in the -25-0 range of latitudes (Fig. 1;
top). Although hundreds of species likely host undiscovered
betacoronaviruses, machine learning predictions have suggested that
these undiscovered reservoirs should follow the same diversity
gradient.\textsuperscript{14}

\begin{figure}
\hypertarget{fig:richness}{%
\centering
\includegraphics{figures/combined_richness.png}
\caption{Top panel: relative diversity of known bat hosts of
betacoronaviruses. This map shows that the region with the largest
number of possible hosts is South-Eastern Asia. Bottom panel: congruence
between the evolutionary distinctiveness of the hosts (grey to blue) and
the viruses (grey to red). By contrast to the richness map, this reveals
that South America has the most evolutionary distinct hosts \emph{and}
viruses, whereas South-Eastern Asia and the Rift Valley region have
mostly distinct viruses. This is congruent with know results about New
World bats being evolutionary distinct, and suggests that they similarly
have distinct viruses.}\label{fig:richness}
}
\end{figure}

Nevertheless, locally diverse and virus-rich bat communities could
represent an increased risk of spillover under climate change through
the creation of novel interactions,\textsuperscript{15} and therefore
the diversity of \emph{Betacoronavirus} strains should similarly be
ccounted for. In fig.~\ref{fig:richness} (bottom), we contrast the
evolutionary distinctiveness of bats and viruses -- this reveals a
slightly different portrait than bat richness alone. Chiropterans can be
classified, from a macro-evolutionary standpoint, as Yangochiroptera and
Yinpterochiroptera elsewhere.\textsuperscript{16,17} Specifically, we
would expect that the so-called ``New World'' group of bats, being more
evolutionary distinct, would also have evolutionary distinct viruses.
Indeed fig.~\ref{fig:richness} (bottom) reveals it to be the case, and
this region harbors a distinct bat-betacoronaviruses complex. This can
be explained by the fact that Yangochiroptera, although not limited to
the western hemisphere, contain the highly diverse adaptive radiation in
the Phyllostomidae,\textsuperscript{18} which is restricted to the
western hemisphere. By contrast, South-Eastern Asia has a lot of
non-evolutionary distinct bats, who nevertheless hosted
evolutionary-distinct viruses.

It is noteworthy that outside of South America, viral evolutionary
distinctiveness does not accurately track host diversity, with some
areas having over-distinct viruses (eastern China but, oddly, not the
rest of southeast Asia). There are a number of likely explanations.
First, given the richness of bats in southeast Asia, many
betacoronaviruses likely remain to be discovered in this region. Indeed,
global predictions highlight that southeast Asia is a likely hostpot of
unconfirmed hosts of betacoronaviruses,\textsuperscript{14} which would
likely result in additional viral discoveries. This idea is unsurprising
given the growing realization, especially since the emergence of
SARS-CoV-2, that unique lineages of similar viruses are widespread in
bats but still mostly undescribed. The most distinct
bats-betacoronaviruses complex is found in South America, a region with
a comparatively lower number of hosts; this matches with isolation
through host vicariance, and may highlight a different co-evolutionary
dynamic. Alternatively, this distinctiveness hostpot may be a product of
under-sampling: South-America is one of the places where the fewest
\emph{Betacoronavirus} sequences have been
discovered,\textsuperscript{2,13,19} resulting in sparser phylogenetic
tree, thereby artificially inflating distinctiveness. Adding more
viruses would bring the distinctiveness of known sequences down.
Finally, South-America is the range of Phyllostomidae, a group of bats
that underwent explosive diversification events,\textsuperscript{18}
which may drive the emergence of multiple viral lineages.

@@

Despite differences in hotspots of evolutionary distinctiveness, there
are reasons to expect that the phylogeography of bats and
betacoronaviruses should show some degree of
congruence.\textsuperscript{20,21} In particular, this should be the
case if viruses can circulate among hosts and co-evolve with local hosts
communities, making their evolutionary process more than a byproduct of
host evolution. High density of hosts sharing the same virus (albeit
possibly different strains) can drive or result from evolution of the
bat antiviral immune system, resulting in spatially distinct
immunological responses, as evidenced in several bat
species.\textsuperscript{22} Immune characteristics that allow bats to
be better adapted to infection by emerging
viruses,\textsuperscript{23,24} in addition to being hardcoded in their
genome,\textsuperscript{25} may be related to a wide variety of
diets,\textsuperscript{22,26} themselves likely to be driven by spatial
effects, especially at the local scale -- bats, indeed, occupy a variety
of environments, and therefore display a variety of adaptations to these
environments.\textsuperscript{27}

@@

In fig.~\ref{fig:biogeo}, we show a projection of the phylogeographic
signal of bats (top) and viruses (bottom) in space; the distinct
groupings (represented by different colors symbolizing positions in the
subspace formed by the first two axes of the PCoA) are essentially
equivalent between the two groups, and can be coarsely delineated as
southeast Asia, Eurasia above a northing of 25, and Africa and south
America. These results suggest that, although the evolutionary
distinctiveness of the bat-betacoronaviruses complex varies spatially,
the system shows an important degree of spatial consistency, with a
reduced number of bioregions. Available information describing the
spillover of zoonotic betacoronaviruses of bat origin where data was
available before and up through the COVID-19 pandemic puts spillover
events of SARS-CoV-2 in Wuhan, China; SARS-CoV in Guangdong, China based
on the presence of closest known viruses circulating in nature, and a
nearby location where serological (antibody) evidence has indicated
human exposure to SARS-like viruses;\textsuperscript{28} MERS-CoV in
Saudi Arabia based on index cases available from a recently-published
compendium of cases.\textsuperscript{29} For the latest event, most if
not all index cases are presumed to be camel-to-human transmission, and
the precise origin point (if it exists) of MERS-CoV in bats is
uncertain. Recent recombinant canine coronavirus spillover events in
Haiti\textsuperscript{30} and Europe\textsuperscript{31} are not
relevant here, as bats' involvement in these cycles of transmission have
been supposed to be non-existent. These index cases fall within
different phylogeographic bioregions (fig.~\ref{fig:biogeo}), which
further highlight the issue that different host-virus sub-systems may
lead to widespread emergence.

@@

From another perspective, areas of high host uniqueness and virus
sharing (red-to-pink) could provide hotspots of \emph{Betacoronavirus}
risk through mixing of unique viruses (via codivergence) and in turn
recombination. Under our framework, such a hotspot was identified in
Madagascar, where most bat species are endemic following evolutionary
divergence from sister species in both African and Asian
continents.\textsuperscript{32} Recent surveillance\textsuperscript{33}
has identified a novel \emph{Betacoronavirus} (in the subgenus
\emph{Nobecovirus}) in Madagascar-endemic pteropid bat species
(\emph{Pteropus rufus}, \emph{Rousettus madagascariensis}), emphasizing
strong proof of principle in model predictions.

@@

Based on the previous result, we extracted the risk component from the
composite map (see Methods), to provide a single measure of risk varying
between 0 and 1. This measure is presented in fig.~\ref{fig:risk}. As
this map represents the potential risk, it must be weighed by the
potential for contacts with humans. As a proxy for this measure, we used
the proportion of build/urban land from the EarthEnv dataset: this is a
reasonable proxy for the density of humans per unit area, which
increases the probability of pathogen spread more
widely.\textsuperscript{34} Since human activity is required to amplify
the frequency of virus encounters and thus create areas of viral
amplification, mapping the potential risk against measures of land use
is required to generate a more actionable assessment of risk. This map
is presented in fig.~\ref{fig:compound}. Most of South America and
Europe are at comparatively lower risk, as although densely populated,
settlements tend to be in areas with lower potential risk. Regions like
Malaysia and the North coast of Australia have a high risk component,
but should represent a relatively lower effective risk due to low human
density. However, this mapping reveals that South-East Asia, the Indian
subcontinent, and parts of sub-Saharan Africa, are at high risk due to
the overlap between built areas and bat communities representing more
opportunities for cross-species transmission of betacoronaviruses. In
looking for the origins of SARS in China,\textsuperscript{35} present
serological evidence that strongest human-animal contact results in
higher risk of virus exposure, regardless of the animal species, but
that different types of contact had different impacts. Ideally,
finer-grained information about human activity (rather than human
presence through anthropisation) could allow to partition this risk
further, alebit at the cost of more hypotheses required to estimate the
amount of risk represented by each activity. Our map of purported high
risk/diversitifcation potential (Madagascar, South-America) overlay with
sampling gaps for \emph{Betacoronavirus},\textsuperscript{36} stressing
the need for spatially targeted monitoring and discovery.

@@

Bats are important reservoir hosts for different classes of
microorganisms, many of which a threat to human
health.\textsuperscript{20,21} Chiropterans emerged around 64 million
years ago and are one of the most diverse mammalian orders, with an
estimated richness of more than 1400 species.\textsuperscript{37,38}
They exhibit a broad variety of habitat use, behaviour, and feeding
strategies, putting them at key positions in the delivery and
provisioning of several ecosystem services, tied to important
ecosystem-derived benefits to human.\textsuperscript{39} For example,
bats are an essential component of many seed-dispersal
networks.\textsuperscript{40} Over two-thirds of bats are know to be
either obligate or facultative insectivores, therefore actively
contributing for agricultural pest control,\textsuperscript{41,42} and
vectors of pathogens that put a risk on human
health.\textsuperscript{43,44} Because bats are globally distributed and
have a long evolutionary history, phylogeographic and biogeographic
approaches are required to shed light on the contemporary distribution
of coevolutionary processes between bats and the pathogens they host.
Not all areas in which bats, viruses, and human are co-occuring are
facing a risk of spillover towards human populations, and the areas in
which this risk exist may not be facing risks of the same nature and
magnitude.

Here, we propose a simple freamework with broad explanatory power that
helps contextualize discoveries like highly divergent nobecoviruses in
Madagascar and the previously-neglected adaptive radiation of
sarbecoviruses outside of southern China and throughout southeast Asia.
In doing so, it advances ecological theory beyond the current state of
the art for global maps of emergence risk. For example, previous studies
that have used host richness as proxy have predicted a high diversity of
unsampled bat viruses,\textsuperscript{13} bat
coronaviruses,\textsuperscript{2} and even specifically
betacoronaviruses\textsuperscript{14} in both the Amazon and southeast
Asia. While we find that both regions are characterized by highly
divergent host and viral communities, our framework identifies key
differences between the regions. We find that Latin America is a hotspot
of both host and viral distinctiveness, suggesting that this branch of
the bat-betacoronavirus complex may be undergoing independent
evolutionary dynamics from the rest of the global pool, but with limited
potential for viral diversification--- a finding that is supported by
previous work indicating a higher rate of codivergence in Latin
America.\textsuperscript{2} In contrast, in southeast Asia, host
richness and viral distinctiveness are high but sharing is low; this
suggests a different type of evolutionary dynamics that could generate
high local diversity of viruses through host switching and viral
recombination (see \emph{e.g.},\textsuperscript{45} as well as the
discovery of recombinant viruses that share genetic material from both
the SARS-CoV and SARS-CoV-2 branches of the Sarbecovirus
lineage).\textsuperscript{46} Both of these regions are priority areas
for sampling, especially given predictions that they contain many bat
hosts of undiscovered betacoronaviruses.\textsuperscript{14,36} However,
both the evolutionary and ecological aspects of emergence risk are
likely higher in southeast Asia---a fact that will only become more
relevant, as bats track shifting climates and exchange viruses with
other species, creating a hotspot of cross-species transmission unique
to the region.\textsuperscript{15}

The diversity and diversification potential of bats responds to
anthropogenic factors others than shifting climates.\textsuperscript{47}
Land use changes could significantly decrease bat suitability, notably
through effects on diet and availability of
habitats.\textsuperscript{48} As our results establish that the
diversification of bats betacoronaviruses happens on top of processes
affecting hosts, biogeographic variation in human population density and
anthropogenic disturbances may feed into co-evolutionary dynamics.
Increase in humans-hosts contacts also increase the risk of emergence of
novel diseases,\textsuperscript{49} so does the changes in landscape
connectivity at local/regional scales.\textsuperscript{50} This
represents a challenge for both conservation strategies and disease
ecology: some areas can a high emergence risk and more potential for the
acquisition of zoonotic viruses through bat-human
encounters.\textsuperscript{51} In particular, the challenge ahead lies
in the need to quantify actual exposure (and risk) accounting for
several transmission scenarios, including both direct and indirect bat -
human interactions, and feeding back into the provision of ecosystem
services by bats.

@@

We downloaded the rangemap of every current bat species that was
classified as an empirically documented host of \emph{Betacoronavirus}
from the previous step, according to recent IUCN
data.\textsuperscript{52} The range maps were subsequently rasterized
using the \texttt{rasterize} function from
\texttt{GDAL}\textsuperscript{53} at a resolution of approximately
100kmx100km. For every pixel in the resulting raster where at least one
bat host of \emph{Betacoronavirus} was present, we extract the species
pool (list of all known bat hosts), which was used to calculate the
following risk assessment components: bat phylogenetic diversity, bat
compositional uniqueness, and predicted viral sharing risk.

@@

For every pixel, we measured Faith's Phylogenetic
Diversity\textsuperscript{54} based on a recent synthetic tree with
robust time calibration, covering about 6000 mammalian
species.\textsuperscript{55} Faith's PD measures the sum of unique
branches from an arbitrary root to a set of tips, and comparatively
larger values indicate a more phylogenetic diverse species pool. We
measured phylogenetic diversity starting from the root of the entire
tree (and not from Chiroptera); this bears no consequences on the
resulting values, since all branches leading up to Chiroptera are only
counted one per species pool, and (as we explain when describing the
assembly of the composite risk map), all individual risk components are
ranged in {[}0,1{]}. This measure incorporates a richness component,
which we chose not to correct for; the interpretation of the
phylogenetic diversity is therefore a weighted species richness, that
accounts for phylogenetic over/under-dispersal in some places.

@@

For every species pool, we measured its Local Contribution to
Beta-Diversity;\textsuperscript{56} LCBD works from a species-data
matrix (traditionally noted as \(\mathbf{Y}\)), where species are rows
and sites are columns, and a value of 1 indicates occurrence. We
extracted the Y matrix assuming that every pixel represents a unique
location, and following best practices\textsuperscript{57} transformed
it using Hellinger's distance to account for unequal bat richness at
different pixels. The correction of raw community data is particularly
important for two reasons: first, it prevents the artifact of richer
sites having higher importance; second, it removes the effect of overall
species richness, which is already incorporated in the phylogenetic
diversity component. High values of LCBD indicate that the pixel has a
community that is on average more dissimilar in species composition than
what is expected knowing the entire matrix, i.e.~a more unique
community. Recent results by\textsuperscript{58} shows that LCBD
measures are robust with regards to spatial scale, and are therefore
applicable at the global scale.

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range overlap across
mammals.\textsuperscript{59} This network stores pairwise values of
viral community similarity. To project viral sharing values into a
single value for every pixel, we averaged the pairwise scores. High
values of the average sharing propensity means that this specific extant
bat assemblage is likely to be proficient at exchanging viruses.

@@

To visualize the aggregated risk at the global scale, we combine the
three individual risk components (phylogenetic diversity, compositional
uniqueness, and viral sharing) using an additive color
model.\textsuperscript{60} In this approach, every risk component gets
assigned a component in the RGB color model (phylogenetic diversity is
green, compositional uniqueness is red, and viral sharing is blue). In
order to achieve a valid RGB measure, all components are re-scaled to
the {[}0,1{]} interval, so that a pixel with no sharing, no phylogenetic
diversity, and no compositional uniqueness is black, and a pixel with
maximal values for each is white. This additive model conveys both the
intensity of the overall risk, but also the nature of the risk as colors
diverge towards combinations of values for three risk components. Out of
the possible combinations, the most risky in terms or rapid
diversification and spillover potential is high phylogenetic diversity
and low viral sharing,\textsuperscript{61} in that this allows multiple
independent host-virus coevolutionary dynamics to take place in the same
location. In the colorimetric space, this correspond to yellow --
because the HSV space is more amenable to calculations for feature
extraction,\textsuperscript{62} we measured the risk level by
calculating the angular distance of the hue of each pixel to a reference
value of 60 (yellow), and weighted this risk level by the value
component. Specifically, given a pixel with colorimetric coordinates
\((h,s,v)\), its ranged weighted risk value is

@@

Subsequently, we tested the hypothesis that the biogeography of bat
betacoronaviruses should track the biogeography of their hosts. To test
this idea, we loosely adapted a method from,\textsuperscript{68,69} who
proposed a phylogenetic method for the delineation of animal
biogeographic regions. In their original method, a distance matrix -
where each row or column represents a geographic raster's grid cell, and
the dissimilarity values are the ``beta diversity similarity'' of their
community assemble - undergoes non-metric multidimensional scaling
(NMDS); the first two axes of the NMDS are projected geographically
using a four-color bivariate map. Here, we build on this idea with an
entirely novel methodology. First, we measure the phylogenetic distance
between the different viruses in the betacoronaviruses tree by using the
cophenetic function in \texttt{ape};\textsuperscript{70} subsequently,
we take a principal components analysis of that distance matrix (readily
interchangeable for NMDS in this case) to project the viral tree into an
n-dimensional space. We then take the first two principal components
and, as with the evolutionary distinctiveness analysis, aggregated these
to a mean host value and projected them using a four-color bivariate
map.

>>>>

\hypertarget{hotspots-of-host-richness-and-viral-diversification-are-distinct}{%
\subsection{Hotspots of host richness and viral diversification are
distinct}\label{hotspots-of-host-richness-and-viral-diversification-are-distinct}}

Bats, the second most diverse groups of mammals, are found worldwide;
gradients in their species richness generally track broader patterns of
mammal diversity, with a striking Neotropical hotspot (especially in the
Amazon basin) and a secondary hotspot centered in the southeast Asian
peninsula. These hotspots of bat diversity are generally presumed to be
hotspots of viral adaptive radiation, and therefore areas of concern for
human health.\textsuperscript{2,13} However, the hotspots of bat
betacoronavirus reservoirs show a distinct pattern, with primary
hotspots (both in terms of size and higher values) of host richness
situated primarily southeast Asia, parts of southern Europe, and to a
lesser extent parts of Africa in the -25-0 range of latitudes
(fig.~\ref{fig:richness}; top). Although hundreds of species likely host
undiscovered betacoronaviruses, machine learning predictions have
suggested that these undiscovered reservoirs should follow the same
diversity gradient.\textsuperscript{14} In principle, these hotspots of
locally-diverse, virus-rich bat communities should drive more adaptive
diversification in their viruses.

\begin{figure}
\hypertarget{fig:richness}{%
\centering
\includegraphics{figures/combined_richness.png}
\caption{Top panel: relative diversity of known bat hosts of
betacoronaviruses. This map shows that the region with the largest
number of possible hosts is South-Eastern Asia. Bottom panel: congruence
between the evolutionary distinctiveness of the hosts (grey to blue) and
the viruses (grey to red).}\label{fig:richness}
}
\end{figure}

However, we find that the global pattern of betacoronavirus phylogenetic
distinctiveness is quite distinct from both bat host richness and
phylogenetic distinctiveness (fig.~\ref{fig:richness}; bottom). In
contrast to the sparsity of Neotropical betacoronavirus hosts, South
America has the most evolutionary distinct hosts \emph{and} viruses,
followed by secondary hotspots in southeast Asia and the Rift Valley
region have mostly distinct viruses. Some degree of sampling bias may
contribute to these patterns: for example, South-America is one of the
places where the fewest bat betacoronavirus sequences have been
generated,\textsuperscript{2,13,15} resulting in a sparser phylogenetic
tree, and artificially inflating distinctiveness; conversely,
disproportionate research effort in eastern China\textsuperscript{16}
may have led to a more complete inventory of the local diversity of
coronaviruses, again inflating these metrics relative to underlying
patterns. Even accounting for these potential biases, though, there are
obvious and distinctive regional regimes of bat-betacoronavirus
diversification.

These global patterns recapitulate the evolutionary history of both the
order Chiroptera and the genus \emph{Betacoronavirus}. Horseshoe bats
(Rhinolophidae) are both the reservoirs of the SARS-like viruses
(subgenus \emph{Sarbecovirus}) and the presumed ancestral host of
\emph{Betacoronavirus}.\textsuperscript{17} The hotspot of host richness
and viral diversity in southeast Asia---both of which are
disproportionately high considering the global landscape of bat species
richness---is almost entirely driven by dynamics in this
clade.\textsuperscript{3,14} In contrast, the Neotropical hotspot is
driven by a combination of distinct bat and virus adaptive radiation.
The hotspot of bat diversity in the Americas is driven by the New World
leaf-nosed bats (Phyllostomidae),\textsuperscript{18} a group of bats
that have undergone multiple explosive diversification events--- but
that are more often hosts of alphacoronaviruses. Out of the four main
groups of betacoronaviruses, only the subgenus \emph{Merbecovirus}---the
outgroup of the betacoronavirus tree---has been found in New World bats;
while found worldwide, the merbecoviruses are almost entirely restricted
to vesper bats (Vespertilionidae) {[}{]}. In this light, the hotspot of
virus distinctiveness in the Americas is the result of isolation by
vicariance--- a unique branch of coronavirus-bat coevolution, separate
from the rest of the globe.

@@

Despite differences in hotspots of evolutionary distinctiveness, there
are reasons to expect that the phylogeography of bats and
betacoronaviruses should show some degree of
congruence.\textsuperscript{19,20} In particular, this should be the
case if viruses can circulate among hosts and co-evolve with local hosts
communities, making their evolutionary process more than a byproduct of
host evolution. High density of hosts sharing the same virus (albeit
possibly different strains) can drive or result from evolution of the
bat antiviral immune system, resulting in spatially distinct
immunological responses, as evidenced in several bat
species.\textsuperscript{21} Immune characteristics that allow bats to
be better adapted to infection by emerging
viruses,\textsuperscript{22,23} in addition to being hardcoded in their
genome,\textsuperscript{24} may be related to a wide variety of
diets,\textsuperscript{21,25} themselves likely to be driven by spatial
effects, especially at the local scale -- bats, indeed, occupy a variety
of environments, and therefore display a variety of adaptations to these
environments.\textsuperscript{26}

@@

In fig.~\ref{fig:biogeo}, we show a projection of the phylogeographic
signal of bats (top) and viruses (bottom) in space; the distinct
groupings (represented by different colors symbolizing positions in the
subspace formed by the first two axes of the PCoA) are essentially
equivalent between the two groups, and can be coarsely delineated as
southeast Asia, Eurasia above a northing of 25, and Africa and south
America. These results suggest that, although the evolutionary
distinctiveness of the bat-betacoronaviruses complex varies spatially,
the system shows an important degree of spatial consistency, with a
reduced number of bioregions. Available information describing the
spillover of zoonotic betacoronaviruses of bat origin where data was
available before and up through the COVID-19 pandemic puts spillover
events of SARS-CoV-2 in Wuhan, China; SARS-CoV in Guangdong, China based
on the presence of closest known viruses circulating in nature, and a
nearby location where serological (antibody) evidence has indicated
human exposure to SARS-like viruses;\textsuperscript{27} MERS-CoV in
Saudi Arabia based on index cases available from a recently-published
compendium of cases.\textsuperscript{28} For the latest event, most if
not all index cases are presumed to be camel-to-human transmission, and
the precise origin point (if it exists) of MERS-CoV in bats is
uncertain. Recent recombinant canine coronavirus spillover events in
Haiti\textsuperscript{29} and Europe\textsuperscript{30} are not
relevant here, as bats' involvement in these cycles of transmission have
been supposed to be non-existent. These index cases fall within
different phylogeographic bioregions (fig.~\ref{fig:biogeo}), which
further highlight the issue that different host-virus sub-systems may
lead to widespread emergence.

@@

From another perspective, areas of high host uniqueness and virus
sharing (red-to-pink) could provide hotspots of \emph{Betacoronavirus}
risk through mixing of unique viruses (via codivergence) and in turn
recombination. Under our framework, such a hotspot was identified in
Madagascar, where most bat species are endemic following evolutionary
divergence from sister species in both African and Asian
continents.\textsuperscript{31} Recent surveillance\textsuperscript{32}
has identified a novel \emph{Betacoronavirus} (in the subgenus
\emph{Nobecovirus}) in Madagascar-endemic pteropid bat species
(\emph{Pteropus rufus}, \emph{Rousettus madagascariensis}), emphasizing
strong proof of principle in model predictions.

@@

Based on the previous result, we extracted the risk component from the
composite map (see Methods), to provide a single measure of risk varying
between 0 and 1. This measure is presented in fig.~\ref{fig:risk}. As
this map represents the potential risk, it must be weighed by the
potential for contacts with humans. As a proxy for this measure, we used
the proportion of build/urban land from the EarthEnv dataset: this is a
reasonable proxy for the density of humans per unit area, which
increases the probability of pathogen spread more
widely.\textsuperscript{33} Since human activity is required to amplify
the frequency of virus encounters and thus create areas of viral
amplification, mapping the potential risk against measures of land use
is required to generate a more actionable assessment of risk. This map
is presented in fig.~\ref{fig:compound}. Most of South America and
Europe are at comparatively lower risk, as although densely populated,
settlements tend to be in areas with lower potential risk. Regions like
Malaysia and the North coast of Australia have a high risk component,
but should represent a relatively lower effective risk due to low human
density. However, this mapping reveals that South-East Asia, the Indian
subcontinent, and parts of sub-Saharan Africa, are at high risk due to
the overlap between built areas and bat communities representing more
opportunities for cross-species transmission of betacoronaviruses. In
looking for the origins of SARS in China,\textsuperscript{34} present
serological evidence that strongest human-animal contact results in
higher risk of virus exposure, regardless of the animal species, but
that different types of contact had different impacts. Ideally,
finer-grained information about human activity (rather than human
presence through anthropisation) could allow to partition this risk
further, alebit at the cost of more hypotheses required to estimate the
amount of risk represented by each activity. Our map of purported high
risk/diversitifcation potential (Madagascar, South-America) overlay with
sampling gaps for \emph{Betacoronavirus},\textsuperscript{16} stressing
the need for spatially targeted monitoring and discovery.

@@

Bats are important reservoir hosts for different classes of
microorganisms, many of which a threat to human
health.\textsuperscript{19,20} Chiropterans emerged around 64 million
years ago and are one of the most diverse mammalian orders, with an
estimated richness of more than 1400 species.\textsuperscript{35,36}
They exhibit a broad variety of habitat use, behaviour, and feeding
strategies, putting them at key positions in the delivery and
provisioning of several ecosystem services, tied to important
ecosystem-derived benefits to human.\textsuperscript{37} For example,
bats are an essential component of many seed-dispersal
networks.\textsuperscript{38} Over two-thirds of bats are know to be
either obligate or facultative insectivores, therefore actively
contributing for agricultural pest control,\textsuperscript{39,40} and
vectors of pathogens that put a risk on human
health.\textsuperscript{41,42} Because bats are globally distributed and
have a long evolutionary history, phylogeographic and biogeographic
approaches are required to shed light on the contemporary distribution
of coevolutionary processes between bats and the pathogens they host.
Not all areas in which bats, viruses, and human are co-occuring are
facing a risk of spillover towards human populations, and the areas in
which this risk exist may not be facing risks of the same nature and
magnitude.

Here, we propose a simple freamework with broad explanatory power that
helps contextualize discoveries like highly divergent nobecoviruses in
Madagascar and the previously-neglected adaptive radiation of
sarbecoviruses outside of southern China and throughout southeast Asia.
In doing so, it advances ecological theory beyond the current state of
the art for global maps of emergence risk. For example, previous studies
that have used host richness as proxy have predicted a high diversity of
unsampled bat viruses,\textsuperscript{13} bat
coronaviruses,\textsuperscript{2} and even specifically
betacoronaviruses\textsuperscript{14} in both the Amazon and southeast
Asia. While we find that both regions are characterized by highly
divergent host and viral communities, our framework identifies key
differences between the regions. We find that Latin America is a hotspot
of both host and viral distinctiveness, suggesting that this branch of
the bat-betacoronavirus complex may be undergoing independent
evolutionary dynamics from the rest of the global pool, but with limited
potential for viral diversification--- a finding that is supported by
previous work indicating a higher rate of codivergence in Latin
America.\textsuperscript{2} In contrast, in southeast Asia, host
richness and viral distinctiveness are high but sharing is low; this
suggests a different type of evolutionary dynamics that could generate
high local diversity of viruses through host switching and viral
recombination (see \emph{e.g.},\textsuperscript{17} as well as the
discovery of recombinant viruses that share genetic material from both
the SARS-CoV and SARS-CoV-2 branches of the Sarbecovirus
lineage).\textsuperscript{43} Both of these regions are priority areas
for sampling, especially given predictions that they contain many bat
hosts of undiscovered betacoronaviruses.\textsuperscript{14,16} However,
both the evolutionary and ecological aspects of emergence risk are
likely higher in southeast Asia---a fact that will only become more
relevant, as bats track shifting climates and exchange viruses with
other species, creating a hotspot of cross-species transmission unique
to the region.\textsuperscript{44}

The diversity and diversification potential of bats responds to
anthropogenic factors others than shifting climates.\textsuperscript{45}
Land use changes could significantly decrease bat suitability, notably
through effects on diet and availability of
habitats.\textsuperscript{46} As our results establish that the
diversification of bats betacoronaviruses happens on top of processes
affecting hosts, biogeographic variation in human population density and
anthropogenic disturbances may feed into co-evolutionary dynamics.
Increase in humans-hosts contacts also increase the risk of emergence of
novel diseases,\textsuperscript{47} so does the changes in landscape
connectivity at local/regional scales.\textsuperscript{48} This
represents a challenge for both conservation strategies and disease
ecology: some areas can a high emergence risk and more potential for the
acquisition of zoonotic viruses through bat-human
encounters.\textsuperscript{49} In particular, the challenge ahead lies
in the need to quantify actual exposure (and risk) accounting for
several transmission scenarios, including both direct and indirect bat -
human interactions, and feeding back into the provision of ecosystem
services by bats.

@@

We downloaded the rangemap of every current bat species that was
classified as an empirically documented host of \emph{Betacoronavirus}
from the previous step, according to recent IUCN
data.\textsuperscript{50} The range maps were subsequently rasterized
using the \texttt{rasterize} function from
\texttt{GDAL}\textsuperscript{51} at a resolution of approximately
100kmx100km. For every pixel in the resulting raster where at least one
bat host of \emph{Betacoronavirus} was present, we extract the species
pool (list of all known bat hosts), which was used to calculate the
following risk assessment components: bat phylogenetic diversity, bat
compositional uniqueness, and predicted viral sharing risk.

@@

For every pixel, we measured Faith's Phylogenetic
Diversity\textsuperscript{52} based on a recent synthetic tree with
robust time calibration, covering about 6000 mammalian
species.\textsuperscript{53} Faith's PD measures the sum of unique
branches from an arbitrary root to a set of tips, and comparatively
larger values indicate a more phylogenetic diverse species pool. We
measured phylogenetic diversity starting from the root of the entire
tree (and not from Chiroptera); this bears no consequences on the
resulting values, since all branches leading up to Chiroptera are only
counted one per species pool, and (as we explain when describing the
assembly of the composite risk map), all individual risk components are
ranged in {[}0,1{]}. This measure incorporates a richness component,
which we chose not to correct for; the interpretation of the
phylogenetic diversity is therefore a weighted species richness, that
accounts for phylogenetic over/under-dispersal in some places.

@@

For every species pool, we measured its Local Contribution to
Beta-Diversity;\textsuperscript{54} LCBD works from a species-data
matrix (traditionally noted as \(\mathbf{Y}\)), where species are rows
and sites are columns, and a value of 1 indicates occurrence. We
extracted the Y matrix assuming that every pixel represents a unique
location, and following best practices\textsuperscript{55} transformed
it using Hellinger's distance to account for unequal bat richness at
different pixels. The correction of raw community data is particularly
important for two reasons: first, it prevents the artifact of richer
sites having higher importance; second, it removes the effect of overall
species richness, which is already incorporated in the phylogenetic
diversity component. High values of LCBD indicate that the pixel has a
community that is on average more dissimilar in species composition than
what is expected knowing the entire matrix, i.e.~a more unique
community. Recent results by\textsuperscript{56} shows that LCBD
measures are robust with regards to spatial scale, and are therefore
applicable at the global scale.

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range overlap across
mammals.\textsuperscript{57} This network stores pairwise values of
viral community similarity. To project viral sharing values into a
single value for every pixel, we averaged the pairwise scores. High
values of the average sharing propensity means that this specific extant
bat assemblage is likely to be proficient at exchanging viruses.

@@

To visualize the aggregated risk at the global scale, we combine the
three individual risk components (phylogenetic diversity, compositional
uniqueness, and viral sharing) using an additive color
model.\textsuperscript{58} In this approach, every risk component gets
assigned a component in the RGB color model (phylogenetic diversity is
green, compositional uniqueness is red, and viral sharing is blue). In
order to achieve a valid RGB measure, all components are re-scaled to
the {[}0,1{]} interval, so that a pixel with no sharing, no phylogenetic
diversity, and no compositional uniqueness is black, and a pixel with
maximal values for each is white. This additive model conveys both the
intensity of the overall risk, but also the nature of the risk as colors
diverge towards combinations of values for three risk components. Out of
the possible combinations, the most risky in terms or rapid
diversification and spillover potential is high phylogenetic diversity
and low viral sharing,\textsuperscript{59} in that this allows multiple
independent host-virus coevolutionary dynamics to take place in the same
location. In the colorimetric space, this correspond to yellow --
because the HSV space is more amenable to calculations for feature
extraction,\textsuperscript{60} we measured the risk level by
calculating the angular distance of the hue of each pixel to a reference
value of 60 (yellow), and weighted this risk level by the value
component. Specifically, given a pixel with colorimetric coordinates
\((h,s,v)\), its ranged weighted risk value is

@@

Subsequently, we tested the hypothesis that the biogeography of bat
betacoronaviruses should track the biogeography of their hosts. To test
this idea, we loosely adapted a method from,\textsuperscript{66,67} who
proposed a phylogenetic method for the delineation of animal
biogeographic regions. In their original method, a distance matrix -
where each row or column represents a geographic raster's grid cell, and
the dissimilarity values are the ``beta diversity similarity'' of their
community assemble - undergoes non-metric multidimensional scaling
(NMDS); the first two axes of the NMDS are projected geographically
using a four-color bivariate map. Here, we build on this idea with an
entirely novel methodology. First, we measure the phylogenetic distance
between the different viruses in the betacoronaviruses tree by using the
cophenetic function in \texttt{ape};\textsuperscript{68} subsequently,
we take a principal components analysis of that distance matrix (readily
interchangeable for NMDS in this case) to project the viral tree into an
n-dimensional space. We then take the first two principal components
and, as with the evolutionary distinctiveness analysis, aggregated these
to a mean host value and projected them using a four-color bivariate
map.
